Sketch Recognition using Particle Swarm Optimizations

### 2009-06-09 trunk/paper2/Paper2.tex
aFter all changes requiredc by doctor

%\begin{abstract}
%%This study 
%Sketches and drawings are widely used as a simple method of expressing thought and designs. The goal of this paper is developing a sketch recognition system that facilitates using sketches in computer systems. The system uses Particle Swarm Optimization (PSO) algorithm to optimally segment the strokes the user draws into meaningful geometric primitives.  These geometric primitives are grouped to formulate symbols which are further identified using a Support Vector Machines (SVM) classifier. Two different segmentation algorithms were evaluated and results show that both algorithms improve the final recognition results. Experiments were conducted on three different dataset, simple presentation (Hs-DB), Electrical Design (EL-DB) and Logic Design (LD-DB) symbols. Results show that the effectiveness of the proposed system on various dataset especially on datasets with low shape diversity as in Logic Design dataset (LD-DS).  % sketch are divided  Users draw symbols and sketchs using a set of pen strokes. 
%%Sketch recognition is defined as the process of identifying symbols that users draw using single or multiple %strokes. Usually, users draw strokes using pens. The sketch recognition system immediately interprets their %strokes as objects that can be easily manipulated. This paper uses Particle Swarm Optimization (PSO) to divide %the strokes the user draws into meaningful geometric primitives. These geometric primitives are grouped to %formulate symbols which are further identified. The results show that using PSO improves segmentation results %which guide the symbol recognition phase. As for recognition we use Support Vector Machines (SVM) classifier %which further improves the final recognition accuracy.  
%\end{abstract}
\pagenumbering{arabic}
\section{Introduction}

@@

\subsection{Segmentation}
\label{seg}
In the segmentation stage, the input stroke is divided into a set of primitives. As shown in Figure \ref{fig:segblock} first an attempt is made to fit the stroke points into a curve or an ellipse using a minimum square error fitting algorithm \cite{ellipsefit}. The ellipse fitting step mainly helps to prevent the system from over segmenting the stroke into multiple lines or curves if the input stroke is an ellipse. If the stroke proved to be an ellipse arc then the segmentation process ends and the system proceeds to the next step. Otherwise, the stroke is passed to two further segmentation algorithms that divide the stroke to either lines or lines and curves. Hence we generate two different segmentations, the system chooses the segmentation with the minimum error. The following sections describe in details the ellipse detection algorithm and the two segmentation algorithms used to divide the stroke. % The input to these algorithms are the stroke points along with the possible dominant points $P_{pd}$ computed during preprocessing. The output is a set of dominant points which are connected by either lines or curves. 
%After two different segmentations are generated, 
%///AllBlockDiagram.jpg  you can change the block to the main block. 
 \begin{figure}
	\centering
		\includegraphics[scale=0.48]{images/blockSmall.jpg}
	\caption{Segmentation Block} 
	\label{fig:segblock}
\end{figure}
\subsection{Ellipse fitting}
%The process starts by computing the center of the stroke bounding box. The bounding box center point is used as the first estimation of the ellipse center. The axes of the ellipse are estimated as $width/2$ and $height/2$ of the stroke bounding box. The least square fitting algorithm is used to minimize the fitting error of the ellipse equation \cite{chernov}.  
This process attempts to fit the stroke points into an ellipse arc; it starts with computing the center of the stroke bounding box. The bounding box center point is used as the first estimation of the center of the ellipse. The axes of the ellipse are estimated as the $width/2$ and $height/2$ of the stroke bounding box where $width$ and $height$ are the width and height of the bonding box. The least square fitting algorithm \cite{chernov} is used to minimize the fitting error of the ellipse Equation (\ref{eq:circleFit})  
\begin{equation}
E = \sum\limits_{i = 0}^N {\frac{{(x_i - x_0 )}}{{a^2 }}^2  + \frac{{(y_i - y_0 )}}{{b^2 }}^2  - 1} 
\label{eq:circleFit}
\end{equation}
where $N$ is number of points in the stroke, $a$ \& $b$ are the length of ellipse axes, $x_0$ \& $y_0$ are the coordinates of the center point and $x_i$ \& $y_i$ are the coordinates of point $i$ in the stroke. A list of new values for $x_0$ , $y_0$ ,$a$ and $b$ are generated randomly from the older values with small increments after each loop.  After few iterations, the final fit error of the estimated ellipse is reported. Another measure is used to compute the efficiency of the final estimated ellipse. Equation \ref{eq:circleError} ensures that the drawn percentage of ellipse is considered. This eliminates fitting a  line into very large ellipse but leaves small ellipses to be fitted as a partial or a full ellipse.

@@

After each iteration of the swarm algorithm (\textsl{AlgS1} and \textsl{AlgS2}), each particle is refined using the following procedures: 
\begin{enumerate}
	\item For each particle $P_i$ each dominant point $P_{ij}$ is checked to find if it was labeled before as a \textit{possible dominant point} $P_{pd}$ (computed as in section \ref{Prepross}). If it was not labeled, the point $P_{ij}$ is moved to the nearest labeled point. This ensures that all of the points generated by the DPSO are possible dominant points $P_{pd}$. 
	\item The particles are tested to make sure that the distance between every two successive dominate point is larger than $min_D$, where $min_D$ is 5\% of the total length of the stroke.  Otherwise, one of these points is removed based on the error cased by its removal.
	\item Each two successive segments are test to determine if they can be merged into a single segment. If merging the two segments have the same or smaller the error, the two segments are merged. This step ensures that the final segmentation of the system has minimum number of segments. 
\end{enumerate}
%1) 2)  % If two points are nearer than $min_D$ then one of the points is removed. 
\subsection{Recognition}
\label{sec:Recognition}
After the user draws all strokes of the symbol, the set of un-recognized strokes is grouped together along with their segmentation as input to the feature extraction process. A composite set of features is used to generate a single feature vector. The following list describes each feature in details.
%\begin{description}
\subsubsection{Structural and geometrical Features(FS1)}
  Features defines the structure of the geometrical symbol.  
%\begin{description}
%	 \item[No. Of segments] Number of segments in the symbol. 
%		\item[No. Of primitives] Number of primitives in the symbol. The feature helps when identifying             symbols with mixed geometric primitives like cylinders and callouts.  
%		\item [No. of curves] Number of curves or ellipses in the symbol. 
%		\item [No. of lines] Number of lines in the symbol. 
%		\item [No. of perpendicular lines] Number of perpendicular lines. 
%		\item [No. of parallel lines] Number of parallel lines. 
%		\item [No. of intersections] Number of intersection between lines and curves. 
%		\item [Size Ratio] Ratio between width to height of the symbol.
%		%\item[No. of holes] Number of holes in the symbols. 
%
%\end{description}
 \begin{itemize}
	 \item \emph{Segments:} Number of segments in the symbol.
	 \item \emph{Strokes:} Number of strokes or partial strokes that created the symbols.  
		\item  \emph{Primitives:} Number of primitives in the symbol. The feature helps when identifying             symbols with mixed geometric primitives like cylinders and callouts.  
		\item \emph{Curves:} Number of curves or ellipses in the symbol. 
		\item \emph{Lines:} Number of lines in the symbol. 
		\item \emph{Perpendicular} lines Number of perpendicular lines. 
		\item \emph{Parallel} lines Number of parallel lines. 
		\item \emph{Intersections} Number of intersection between lines and curves. 
		\item \emph{T intersections} Number of T intersections. 
		\item \emph{L intersections} Number of L intersections. 
		\item \emph{X intersections} Number of X intersections.

		%\item[No. of holes] Number of holes in the symbols. 
\end{itemize}
\subsubsection{Rubine Feature Set (FS2)}
  Features introduced by Rubine\cite{gestureexample12} for single stroke gestures. Feature such as length of stroke, smoothness of symbol and properties of the bounding box of the symbol.

\subsubsection{Statistical Features (FS3)}  
\begin{itemize}
	\item \emph{Zernike moments } Zernike moments of order n (10 or 20).  \cite{HeloiseBeautification}. 
\end{itemize}
\subsubsection{Composite Features (FS4)}
 Features that are composites of different geometrical and statistical symbol. 
	\begin{itemize}
\item \emph{Size Ratio} Ratio between width to height of the symbol.
	\item \emph{Ink density} Compute the density of points inside its bounding box\cite{GeometryAndDomain102}.   
 	\item \emph{Convex Hull Area} Area of convex hull with respect to area of bounding box of symbol.
	\item \emph{Convex Hull Perimeter} Perimeter of convex hull with respect to total length of symbol.
		\item \emph{Mean Centroidal radius} The Mean of the centroidal radius which is the distance from each point in the symbol to the center of gravity.

	\item \emph{Mean Time difference} The mean of the time difference between each two successive points in the symbol. %Different strokes are appended to construct a single path.
	%\item [Ra]
  \end{itemize}

%\end{description}
%The features used consist of Rubine feature set,  of order 21 , ink density as well as some structural and spatial information like number of perpendiculars lines ,number of parallel lines and number of different types of primitives in each symbol. 
After computing the features the symbol is introduced to the classifier. The system uses Support Vector Machine (SVM) classifier with Linear kernel \cite{libsvm}. An OVO classifier structure (object versus object) is used to handle multi-class.% A validation algorithm is used to choose the Kernel variables parameters $c, \gamma$.

\section{Experimental}
\label{sec:Experiments}
In our experiment we used three different datasets. First dataset was collected by Hse and Newton\cite{HeloiseBeautification} (Hs-DB). The data are drawn by 16 users each of them was asked to draw 13 shapes from 30 to 50 times. Figure \ref{fig:symbolSet} shows a set of the shapes used in the data set. We collected the second and third datasets. Figure \ref{fig:ELsymbolSet} and \ref{fig:LogicsymbolSet} shows the symbols gathered from 7 different users in electrical (EL-DB) and logic design (LD-DB) domain respectively.  Table \ref{tab:datasets} shows a  comparison between the three datasets used in our experiments. The datasets were divided into training set and test set. Five different splits  for the training and test data are generated from each dataset. The results displayed are the average recognition accuracy of the five splits. The accuracy is computed as the number of correctly recognized samples divided by the total number of samples in the test.

\begin{figure}
\centering 
\label{fig:symbolSet}
\fbox{ \parbox{5cm}{% 
		\includegraphics[scale=0.5]{images/symbolSet.PNG}	}}
		\caption{The Hs-DB Symbol Set}.
\end{figure}
\begin{figure}
\centering 
\label{fig:LogicsymbolSet}

		\includegraphics[scale=0.5]{images/logicSet.jpg}	 
		\caption{The Digital Design Symbols Dataset(LD-DB)} Symbols used in the dataset.
\end{figure}
\begin{figure}
\centering 
\label{fig:ELsymbolSet}
\fbox{ \parbox{4.5cm}{% 
		\includegraphics[scale=0.5]{images/EelectImage.jpg}	}}
		\caption{The Electrical Symbols Dataset(EL-DB)} .
\end{figure}

\begin{table}
\begin{center}
\scalebox{0.9}{
 \begin{tabular}{|p{2cm}|p{1.5cm}|p{1.5cm}|p{1.5cm}|}
 \hline
 & Hs-DB &  EL-DB & LD-DB \\ \hline
No of samples  &  7791 & 2764 & 1859 \\ \hline
No of categories  & 13 & 15 & 8 \\ \hline
Avg. Samples per Category& 600 & 184 & 232 \\\hline 
No of users  & 16 & 7 & 7 \\  \hline
Balanced & Yes&No & No \\ \hline
Splits  & 5 random splits  & 5 random splits  & 5 random splits  \\ \hline
Examples & Figure \ref{fig:symbolSet}  & Figure \ref{fig:ELsymbolSet}  & Figure \ref{fig:LogicsymbolSet}   \\ \hline
\end{tabular}
}
\caption[Datasets Comparisons]{Datasets Comparisons} 
\label{tab:datasets}

@@

\subsection{Algorithms comparisons}
\label{sec:AlgExp}
A set of tests were performed on the system to enhance non ellipse fitting algorithm. The effect of number of agents, the error thresholds and various other parameters were investigated. The result of these tests are shown in Figure \ref{fig:swarmtesting} and   \ref{fig:swarmtesting2}. Figure \ref{fig:swarmtesting} shows that as the number of agents increases the error value decrease (error of the segmentation algorithm). The final number of agents and maximum swarm iteration used in the system was based on a trade off between error achieved versus the computation time. Other algorithm parameters such as $c_1$,$c_2$ that were mentioned in Section\ref{sec:ParticleSwarmAlgorithm} are tested using similar tests (Figure \ref{fig:swarmtesting} and \ref{fig:swarmtesting2}).

 \begin{figure}
	\centering		
	 \includegraphics[scale=0.3]{images/swarmAgentTest.jpg}
	 	\caption{Swarm testing} Segmentation Error Vs. Number of agents
	 	\label{fig:swarmtesting}
	%\caption{Experiments results :} a)    b) c) 
\end{figure}

\begin{figure}
	\centering		
	 \includegraphics[scale=0.4]{images/swarmParamterC.jpg}
	 	\caption{Swarm testing} Segmentation Error Vs. PSO parameters.
	 	\label{fig:swarmtesting2}
	%\caption{Experiments results :} a)    b) c)

\end{figure}

We performed several experiments to evaluate the presented recognition system. Firstly we tested recognition accuracy of shapes in the data set with both algorithms (\textsl{AlgS1} and \textsl{AlgS2}). We also implemented the segmentation algorithm described in \cite{earlyprocess} (\textsl{Alg3}) to use as reference to compare it with our swarm algorithms. Figure \ref{fig:test1} and \ref{fig:testLD} shows the accuracy achieved by each algorithm. The two swarm algorithms were tested with and without the ellipse fitting module. The ellipse detection module appears to be superior. The result shows that combining both  (\textsl{AlgS1} and \textsl{AlgS2}) out perform any single algorithm. %show that combining algoirthmsachieve higher accuracy than any single algoirthm. 
%improves the results.% with both the \textit{DPSO} algorithms.% The results show that both \textit{DPSO} algorithms achieve better result than other algorithms.

 \begin{figure}
	\centering		
	 \includegraphics[scale=0.4]{images/testAlg.jpg}
	 	\caption{Algorithm comparison on HS-DB} The recognition rate of different algorithms. 
	 	\label{fig:test1}
	%\caption{Experiments results :} a)    b) c) 
\end{figure}

\begin{figure}
	\centering		
	 \includegraphics[scale=0.4]{images/LDAlg.jpg}
	 	\caption{Algorithm comparison on LD-DB} The recognition rate of different algorithms. 
	 	\label{fig:testLD}
	%\caption{Experiments results :} a)    b) c) 
\end{figure}

\begin{figure}
	\centering		
	 \includegraphics[scale=0.4]{images/AlgEL.jpg}
	 	\caption{Algorithm comparison on EL-DB} The recognition rate of different algorithms. 
	 	\label{fig:testEL}
	%\caption{Experiments results :} a)    b) c) 
\end{figure}

\subsection{Shape Complexity Analysis}
\label{sec:ShapeComplexityExperiments}

\begin{figure*}
	\centering
		\includegraphics[scale=0.5]{images/testsym.jpg}
	\caption{Symbols Comparison on HS-DB} The effect of symbol complexity.  %The graph shows the recognition rate of each symbol using different algorithms. 
	\label{fig:test2}
\end{figure*}

>>>>

%\begin{abstract}
%%This study 
%Sketches and drawings are widely used as a simple method of expressing thought and designs. The goal of this paper is developing a sketch recognition system that facilitates using sketches in computer systems. The system uses Particle Swarm Optimization (PSO) algorithm to optimally segment the strokes the user draws into meaningful geometric primitives.  These geometric primitives are grouped to formulate symbols which are further identified using a Support Vector Machines (SVM) classifier. Two different segmentation algorithms were evaluated and results show that both algorithms improve the final recognition results. Experiments were conducted on three different dataset, simple presentation (Hs-DS), Electrical Design (EL-DS) and Logic Design (LD-DS) symbols. Results show that the effectiveness of the proposed system on various dataset especially on datasets with low shape diversity as in Logic Design dataset (LD-DS).  % sketch are divided Users draw symbols and sketches using a set of pen strokes. 
%%Sketch recognition is defined as the process of identifying symbols that users draw using single or multiple %strokes. Usually, users draw strokes using pens. The sketch recognition system immediately interprets their %strokes as objects that can be easily manipulated. This paper uses Particle Swarm Optimization (PSO) to divide %the strokes the user draws into meaningful geometric primitives. These geometric primitives are grouped to %formulate symbols which are further identified. The results show that using PSO improves segmentation results %which guide the symbol recognition phase. As for recognition we use Support Vector Machines (SVM) classifier %which further improves the final recognition accuracy.  
%\end{abstract}
\pagenumbering{arabic}
\section{Introduction}

@@

\subsection{Segmentation}
\label{seg}
In the segmentation stage, the input stroke is divided into a set of primitives. As shown in Figure \ref{fig:segblock} first an attempt is made to fit the stroke points into a curve or an ellipse using a minimum square error fitting algorithm \cite{ellipsefit}. The ellipse fitting step mainly helps to prevent the system from over segmenting the stroke into multiple lines or curves if the input stroke is an ellipse. If the stroke proved to be an ellipse arc then the segmentation process ends and the system proceeds to the next step. Otherwise, the stroke is passed to two further segmentation algorithms that divide the stroke to either lines or lines and curves. Hence we generate two different segmentations; the system chooses the segmentation with the minimum error. The following sections describe in details the ellipse detection algorithm and the two segmentation algorithms used to divide the stroke. % The input to these algorithms are the stroke points along with the possible dominant points $P_{pd}$ computed during preprocessing. The output is a set of dominant points which are connected by either lines or curves. 
%After two different segmentations are generated, 
%///AllBlockDiagram.jpg  you can change the block to the main block. flowchart
 \begin{figure}
	\centering
		\includegraphics[scale=0.48]{images/flowchart.jpg}
	\caption{Segmentation process flow chart} Inputs to the segmentation proces are the input stroke and \textit{possible dominant points} $P_{pd}$. The output is either the stroke fitted as an ellipse or the best segmentation generated from the two segmentation algorithms.  % First, attempts to fit stroke as an ellipse. Then two different segmentation are generated using (AlgS1 and AlgS2). Finaly, the Minimum error of segmentation as the best segmentation. 
	\label{fig:segblock}
\end{figure}

% \begin{figure}
%	\centering
%		\includegraphics[scale=0.48]{images/blockSmall.jpg}
%	\caption{Segmentation Block} 
%	\label{fig:segblock}
%\end{figure}

\subsection{Ellipse fitting}
%The process starts by computing the center of the stroke bounding box. The bounding box center point is used as the first estimation of the ellipse center. The axes of the ellipse are estimated as $width/2$ and $height/2$ of the stroke bounding box. The least square fitting algorithm is used to minimize the fitting error of the ellipse equation \cite{chernov}.  
This process attempts to fit the stroke points into an ellipse arc; it starts with computing the center of the stroke bounding box. The bounding box center point is used as the first estimation of the center of the ellipse. The axes of the ellipse are estimated as the $width/2$ and $height/2$ of the stroke bounding box where $width$ and $height$ are the width and height of the bonding box. The least square fitting algorithm \cite{chernov} is used to minimize the fitting error of the ellipse Equation (\ref{eq:circleFit})  
\begin{equation}
E = \sum\limits_{i = 0}^N {\frac{{(x_i - x_0 )}}{{a^2 }}^2  + \frac{{(y_i - y_0 )}}{{b^2 }}^2  - 1} 
\label{eq:circleFit}
\end{equation}
where $N$ is number of points in the stroke, $a$ \& $b$ are the length of ellipse axes, $x_0$ \& $y_0$ are the coordinates of the center point and $x_i$ \& $y_i$ are the coordinates of point $i$ in the stroke. A list of new values for $x_0$ , $y_0$ ,$a$ and $b$ are generated randomly from the older values with small increments after each loop.  After few iterations, the final fit error of the estimated ellipse is reported. Another measure is used to compute the efficiency of the final estimated ellipse. Equation \ref{eq:circleError} ensures that the drawn percentage of ellipse is considered. This eliminates fitting a line into very large ellipse but leaves small ellipses to be fitted as a partial or a full ellipse.

@@

After each iteration of the swarm algorithm (\textsl{AlgS1} and \textsl{AlgS2}), each particle is refined using the following procedures: 
\begin{enumerate}
	\item For each particle $P_i$ each dominant point $P_{ij}$ is checked to find if it was labeled before as a \textit{possible dominant point} $P_{pd}$ (computed as in section \ref{Prepross}). If it was not labeled, the point $P_{ij}$ is moved to the nearest labeled point. This ensures that all of the points generated by the DPSO are possible dominant points $P_{pd}$. 
	\item The particles are tested to make sure that the distance between every two successive dominate point is larger than $min_D$, where $min_D$ is 5\% of the total length of the stroke.  Otherwise, two temperary particles are evaluated by removing either point from the current particle. After that, one of these points is removed based on the error caused by its removal.
	\item Each two successive segments are tested to determine if they can be merged into a single segment. If merging the two segments results in segmentation that has an error value less than or equal to the error before merging, the two segments are merged. This step ensures that the final segmentation of the system has minimum number of segments. 
	% the same or smaller the error
\end{enumerate}
%1) 2)  % If two points are nearer than $min_D$ then one of the points is removed. 
\subsection{Recognition}
\label{sec:Recognition}
After the user draws all strokes of the symbol, the set of un-recognized strokes is grouped together along with their segmentation as input to the feature extraction process. A composite set of features is used to generate a single feature vector. The following list describes each feature in details.
%\begin{description}
\subsubsection{Structural and geometrical Features (FS1)}
  They are features defines the structure of the geometrical symbol such as;   
%\begin{description}
%	 \item[No. Of segments] Number of segments in the symbol. 
%		\item[No. Of primitives] Number of primitives in the symbol. The feature helps when identifying             symbols with mixed geometric primitives like cylinders and callouts.  
%		\item [No. of curves] Number of curves or ellipses in the symbol. 
%		\item [No. of lines] Number of lines in the symbol. 
%		\item [No. of perpendicular lines] Number of perpendicular lines. 
%		\item [No. of parallel lines] Number of parallel lines. 
%		\item [No. of intersections] Number of intersection between lines and curves. 
%		\item [Size Ratio] Ratio between width to height of the symbol.
%		%\item[No. of holes] Number of holes in the symbols. 
%
%\end{description}
 \begin{itemize}
	 \item \emph{Segments:} The number of segments in the symbol.
	 \item \emph{Strokes:} The number of strokes or partial strokes that created the symbols.  
		\item  \emph{Primitives:} The number of primitives in the symbol. This feature helps when identifying  symbols with mixed geometric primitives such as cylinders and callouts.  
		\item \emph{Curves:} The number of curves or ellipses in the symbol. 
		\item \emph{Lines:} The number of lines in the symbol. 
		\item \emph{Perpendicular}\emph{lines:} The number of perpendicular lines. 
		\item \emph{Parallel}\emph{lines:} The number of parallel lines. 
		\item \emph{Intersections:} The number of intersection between lines and curves. 
		\item \emph{T intersections:} The number of T intersections. 
		\item \emph{L intersections:} The number of L intersections. 
		\item \emph{X intersections:} The number of X intersections.	
		%\item[No. of holes] Number of holes in the symbols. 
\end{itemize}
\subsubsection{Rubine Features (FS2)}
  These features introduced by Rubine \cite{gestureexample12} for single stroke gestures. Such as length of stroke, cosine of starting angle, smoothness of symbol and the properties of the bounding box of the symbol.

\subsubsection{Statistical Features (FS3)} 
These features use moments and statistical properties such as;
\begin{itemize}
	\item \emph{Zernike moments:} The zernike moments are orthogonal moments which define the distribution of point in the input space. They are invariant to both reflection and rotation of the input shape \cite{ZerMomentOrthogonal}. %The zernike moment were used in \cite{HeloiseBeautification} t. 
\end{itemize}
\subsubsection{Composite Features (FS4)}
 These features are composites of different geometrical and statistical properties of the symbol. 
	\begin{itemize}
\item \emph{Size Ratio:} The ratio between width and height of the symbol.
	\item \emph{Ink density:} The density of points inside its bounding box\cite{GeometryAndDomain102}.   
 	\item \emph{Convex Hull Area:} The area of convex hull with respect to area of bounding box of symbol.
	\item \emph{Convex Hull Perimeter:} The perimeter of convex hull with respect to total length of symbol.
		\item \emph{Mean Centroidal radius:} The Mean of the centroidal radius is the distance from each point in the symbol to the center of gravity.
	\item \emph{Mean Time difference:} The mean of the time difference between each two successive points in the symbol. %Different strokes are appended to construct a single path.
	%\item [Ra]
  \end{itemize}  
%\end{description}
%The features used consist of Rubine feature set,  of order 21 , ink density as well as some structural and spatial information like number of perpendiculars lines ,number of parallel lines and number of different types of primitives in each symbol. 
After computing the features, the symbol is introduced to the classifier as a feature vector. The system uses Support Vector Machine (SVM) classifier with Linear kernel \cite{libsvm}. An OVO classifier structure (object versus object) is used to handle multi-class problem.% A validation algorithm is used to choose the Kernel variables parameters $c, \gamma$.

\section{Experimental}
\label{sec:Experiments} 
%Figure \ref{fig:symbolSet} shows a set of the shapes used in this data set. 
In our experiment we used three different datasets. The first dataset (Figure \ref{fig:HsSet}) was collected by Hse and Newton\cite{HeloiseBeautification} (Hs-DS). The data are drawn by 16 users each of them was asked to draw 13 shapes from 30 to 50 times. Figure \ref{fig:ELsymbolSet} and \ref{fig:LogicsymbolSet} shows the symbols we manually gathered from 7 different users in electrical (EL-DS) and logic design (LD-DS) domain respectively.  Table \ref{tab:datasets} shows a comparison between the three datasets used in our experiments. The datasets were divided into training and test sets. Five different splits for the training and test data are generated from each dataset. The results displayed are the average recognition accuracy of the five splits. The accuracy is computed as the number of correctly recognized samples divided by the total number of samples in the test. 
\begin{figure}
\centering

\fbox{ \parbox{5cm}{% 
		\includegraphics[scale=0.5]{images/symbolSet.PNG}	}}
		\caption{The Hs-DS Symbol Set} The dataset consist of simple presentation symbols.
		\label{fig:HsSet} 
\end{figure}

 \begin{figure}
			\subfigure[The Digital Design Symbols Dataset(LD-DS)]{\label{fig:LogicsymbolSet} \includegraphics[scale=0.5]{images/logicSet.jpg}}
			\hfill
			\subfigure[The Electrical Symbols Dataset(EL-DS)]
			 {\label{fig:ELsymbolSet}\includegraphics[scale=0.45]{images/EelectImage.jpg}}
%			 \label{fig:ELsymbolSet} 
%			 \fbox{ 
%			 \parbox{4cm}{\includegraphics[scale=0.5]{images/EelectImage.jpg}	}}}

	\caption{The datasets that were manually collected}  a)  The (LD-DS) dataset consist of symbols that are visually similar (i.e. has low shape diversity). b) The (EL-DS) dataset consist of complex topological symbols. 
\end{figure}

%\begin{figure}
%\centering 
%\label{fig:LogicsymbolSet}
% 
%		\includegraphics[scale=0.5]{images/logicSet.jpg}	 
%		\caption{The Digital Design Symbols Dataset(LD-DS)} The dataset consist of symbols that are visually simillar (i.e. has low shape diversity). 
%\end{figure}
%\begin{figure}
%\centering 
%\label{fig:ELsymbolSet}
%\fbox{ \parbox{4.5cm}{% 
%		\includegraphics[scale=0.5]{images/EelectImage.jpg}	}}
%		\caption{The Electrical Symbols Dataset(EL-DS)} . The dataset consist of complex toplogical symbols.
%\end{figure}

\begin{table}
\begin{center}
\scalebox{0.8}{
 \begin{tabular}{|p{3cm}|p{1.5cm}|p{1.5cm}|p{1.5cm}|}
 \hline
 & Hs-DS &  EL-DS & LD-DS \\ \hline
No of samples  &  7791 & 2764 & 1859 \\ \hline
No of categories  & 13 & 15 & 8 \\ \hline
Avg. Samples per Category& 600 & 184 & 232 \\\hline 
No of users  & 16 & 7 & 7 \\  \hline
Balanced & Yes&No & No \\ \hline
Splits  & 5 random splits  & 5 random splits  & 5 random splits  \\ \hline
Examples & Figure \ref{fig:HsSet}  & Figure \ref{fig:ELsymbolSet}  & Figure \ref{fig:LogicsymbolSet}   \\ \hline
\end{tabular}
}
\caption[Datasets Comparisons]{Table compares between datasets used in experiments in terms of number of samples, number of samples per category, etc...} 
\label{tab:datasets}

@@

A set of tests were performed on the system to enhance non-ellipse fitting algorithms. The effect of the number of particles, the error thresholds and various other parameters were investigated. The results of these tests are shown in Figures \ref{fig:swarmtesting} and   \ref{fig:swarmtesting2}. Figure \ref{fig:swarmtesting} shows that as the number of agents increases the error value decrease (error of the segmentation algorithm). The final number of particles and maximum swarm iterations used in the system was based on a tradeoff between error achieved versus the computation time. The effect of other algorithm parameters such as $c_1$, $c_2$ (mentioned in Section\ref{sec:ParticleSwarmAlgorithm}) is tested using similar tests (Figure \ref{fig:swarmtesting}, \ref{fig:swarmtesting2}).

 \begin{figure}
	\centering		
	 	\subfigure[Segmentation Error Vs. Number of particles]{	\label{fig:swarmtesting} \includegraphics[scale=0.3]{images/swarmIterationVsParticles.jpg}}
	 %\includegraphics[scale=0.3]{images/swarmAgentTest.jpg}} 	\caption{}

	%\caption{Experiments results :} a)    b) c) 
	%\end{figure}

%\begin{figure}
	%\centering

		\subfigure[Segmentation Error Vs. PSO parameters.]{	\label{fig:swarmtesting2}  \includegraphics[scale=0.4]{images/swarmParamterC.jpg}}

	 	\caption{}a) The error decrease as the number of swarm iteration increase until it reach a saturation.   b) The behavior of segmentation error versus c1.

	%\Segmentation Error Vs. PSO parameters. caption{Experiments results :} a)    b) c)

\end{figure} 
\subsection{Algorithms comparisons}
\label{sec:AlgExp}
We performed several experiments to evaluate the presented recognition system. Firstly, we tested recognition accuracy of shapes in the data set with both algorithms (\textsl{AlgS1} and \textsl{AlgS2}). We also implemented the segmentation algorithm described in \cite{earlyprocess} (\textsl{Alg3}) to use as reference to compare it with our swarm algorithms. Figure \ref{fig:test1} and \ref{fig:testLD} shows the accuracy achieved by each algorithm. The two swarm algorithms were tested with and without the ellipse fitting module. Combining the ellipse detection module improves the performance. The result shows that (\textsl{AlgS2}) gives better performance than Alg3 and AlgS1 in all datasets.  The results also shows that combining (\textsl{AlgS1} and \textsl{AlgS2} using minimum error) outperforms AlgS2 in Hs-Ds only. This behavior is explained by noticing that features are computed based on the chosen segmentation. This means that different segmentation for a symbol generates different set of features used for recognition. This behavior affect Ls-Ds and EL-Ds more than Hs-Ds due to the higher topological and structure symbols in both datasets than of Hs-Ds.  %   can be explained by the simplicity of the symbols in Hs-DS % and LD-Ds unlike EL-DS where AlgS2 is the best algorithm. This behavior is explained by the structure of the symbols in each dataset. El-DS consist of highly topological and structure symbols that are usually drawn using multiple small strokes. This leads to    %   out perform any single algorithm in HS-DS and LD-DS.  %show that combining algorithms achieve higher accuracy than any single algorithm. 
%improves the results.% with both the \textit{DPSO} algorithms.% The results show that both \textit{DPSO} algorithms achieve better result than other algorithms.

 \begin{figure}[h]
	\centering		
	\subfigure[Algorithm comparison on HS-DS]{\label{fig:test1} \includegraphics[scale=0.4]{images/testAlg.jpg}}
	 %	\caption{} The recognition rate of different algorithms.

	%\caption{Experiments results :} a)    b) c) 
%\end{figure}

%\begin{figure}[h]
%	\centering		
		\subfigure[Algorithm comparison on LD-DS]{ 	 	\label{fig:testLD} \includegraphics[scale=0.4]{images/LDAlg.jpg}}
	% 	\caption{} The recognition rate of different algorithms.

	%\caption{Experiments results :} a)    b) c) 
%\end{figure}

%\begin{figure}[h]
%	\centering		
		\subfigure[Algorithm comparison on EL-DS]{	 	\label{fig:testEL}  \includegraphics[scale=0.4]{images/AlgE.jpg}}
	 	\caption{ The recognition rate of different algorithms. }

	%\caption{Experiments results :} a)    b) c) 
\end{figure}

\subsection{Shape Complexity Analysis}
\label{sec:ShapeComplexityExperiments}
\begin{figure*}
	\centering 
		\subfigure[Symbols Comparison on HS-DS]{\label{fig:test2}
		\includegraphics[scale=0.35]{images/testsym.jpg}}
	%\caption{} The effect of symbol complexity.  %The graph shows the recognition rate of each symbol using different algorithms. 
%\end{figure*}  
%\begin{figure*}
%	\centering
\subfigure[Symbols Comparison on LD-DS symbols]{\label{fig:LDtest2}
		\includegraphics[scale=0.5]{images/DLSymbols.jpg}}
	%\caption{}.  %The graph shows the recognition rate of each symbol using different algorithms. 
	\caption {Symbol Comparison} The accuracy of achieved by each algorithm in each symbol. a) Low accuracy level appears in triangle, heart, and moon symbols. b) Low accuracy level appears only in symbols that are visually confusing (i.e Nor, OR, Nand gates). 
\end{figure*}  
The second experiment we implemented was to investigate the effect of symbol complexity and its type on the recognition rate. Figure \ref{fig:test2} and \ref{fig:LDtest2} show the achieved accuracy of each symbol by our algorithms compared to (\textsl{Alg3}) \cite{earlyprocess}. It is clearly noted that symbols that have only line segments achieve higher accuracy rate than other symbols (such as square, parallegram). The results also indicate that algorithm \textsl{AlgS1} achieves better performance than algorithm \textsl{AlgS2} in the symbols that consists only of lines. This is understandable because algorithm \textsl{AlgS1} divides strokes into line segments only but \textsl{AlgS2} is able to divide strokes into lines and curves based on the minimum error of the segment itself. Figure \ref{fig:LDtest2} shows that in logic design (LD-DS) \textsl{AlgS2} achieved better performance than \textsl{Alg3} and  \textsl{AlgS1}.  This proves that for complex shapes that are combination of lines and arcs are \textsl{AlgS2} achieves better segmentation. Algorithm \textsl{Alg3} gives good performance as long as the symbols consist of combination of lines and curves, if the stroke consists of curve or lines only the algorithm may lead to wrong segmentation result. This is because the system divides the stroke first into line segments then tries to decide whether each segment can be represented better as a curve. However, in \textsl{AlgS2} the curve segments are tested while choosing the best segmentation. Combining both algorithm \textsl{AlgS1} and \textsl{AlgS2} improved the recognition rate of all symbols. The penalty for this improved performance is the computational time required to run both swarm algorithms. Table \ref{tab:ConfusionMatrix} shows the confusion matrix for the (HS-DS) symbols. The table shows that an error occurs mostly between two sets of symbols: moon with triangles and trapezoid with triangles. This observation is understandable as the symbols are visually similar but the system must be able to differentiate between them. This leads us to the next experiment to choose the best set of feature to recognize symbols. 
\begin{table*}
	\centering
	%\small
%	\scalebox{0.7}{
		%	\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|c|c|c|c|}\hline 
 \include{conf}
 		%\end{tabular}
 	%}
	\caption{Confusion Matrix of Hs-DS}
	\label{tab:ConfusionMatrix}
	%\end{minipage}
\end{table*}
% \begin{table*}
% 	\centering
% 	%\small
%	\scalebox{0.7}{
% 		%	\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|c|c|c|c|}\hline 
%  \include{conf2}
%  		%\end{tabular}
%  	%}
% 	\caption{Confusion Matrix of DL-DS}
% 	\label{tab:ConfusionMatrix}
% 	%\end{minipage}
% \end{table*}
\subsection{Features Analysis}
